---
title: "Homework 2"
author: "Halenur Kocaman - 2015402036"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Libraries

The libraries used for the homework are as follows:

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(readr)
library(dplyr)
library(tidyverse)
library(scatterplot3d)
library(ggplot2)

```

##Reading the Data

```{r}
x<-read.table("C:/Users/asus/Desktop/homework 2/UWave/uWaveGestureLibrary_X_TRAIN.txt")
y<-read.table("C:/Users/asus/Desktop/homework 2/UWave/uWaveGestureLibrary_Y_TRAIN.txt")
z<-read.table("C:/Users/asus/Desktop/homework 2/UWave/uWaveGestureLibrary_Z_TRAIN.txt")

colnames(x)[1]<-"class"
colnames(y)[1]<-"class"
colnames(z)[1]<-"class"
```


##TASK 1

First task aims to visualize the position obtained from accelaration information. Then, it will be checked whether they are consistent with the defined class shape or not. Color change from gray to black is made  to understand the differences stemming from directions. According to the 3D scatter plots below, shapes are consistent with the defined classes.

###Class 1

```{r}
ins_1<- as.data.frame(cbind(x=t(filter(x,class==1))[,1],y=t(filter(y,class==1))[,1],z=t(filter(z,class==1))[,1]))[-1,]

ins_1<- data.frame(ins_1,vel_x=cumsum(ins_1$x),vel_y=cumsum(ins_1$y),vel_z=cumsum(ins_1$z))
ins_1<- data.frame(ins_1,pos_x=cumsum(ins_1$vel_x),pos_y=cumsum(ins_1$vel_y),pos_z=cumsum(ins_1$vel_z))
rownames(ins_1) <- 1:nrow(ins_1)
ins_1$row<-as.numeric(rownames(ins_1))/nrow(ins_1)
scatterplot3d(ins_1[,7:9],color=gray(ins_1$row),xlab = "X",ylab="Y",zlab="Z")

```

###Class 2

```{r}
ins_2<- as.data.frame(cbind(x=t(filter(x,class==2))[,2],y=t(filter(y,class==2))[,2],z=t(filter(z,class==2))[,2]))[-1,]

ins_2<- data.frame(ins_2,vel_x=cumsum(ins_2$x),vel_y=cumsum(ins_2$y),vel_z=cumsum(ins_2$z))
ins_2<- data.frame(ins_2,pos_x=cumsum(ins_2$vel_x),pos_y=cumsum(ins_2$vel_y),pos_z=cumsum(ins_2$vel_z))

rownames(ins_2) <- 1:nrow(ins_2)
ins_2$row<-as.numeric(rownames(ins_2))/nrow(ins_2)
scatterplot3d(ins_2[,7:9],color=gray(ins_2$row),xlab = "X",ylab="Y",zlab="Z")
#scatterplot3d(ins_2[,7:9],color="red")

```

###Class 3

```{r}
ins_3<- as.data.frame(cbind(x=t(filter(x,class==3))[,2],y=t(filter(y,class==3))[,2],z=t(filter(z,class==3))[,2]))[-1,]

ins_3<- data.frame(ins_3,vel_x=cumsum(ins_3$x),vel_y=cumsum(ins_3$y),vel_z=cumsum(ins_3$z))
ins_3<- data.frame(ins_3,pos_x=cumsum(ins_3$vel_x),pos_y=cumsum(ins_3$vel_y),pos_z=cumsum(ins_3$vel_z))

rownames(ins_3) <- 1:nrow(ins_3)
ins_3$row<-as.numeric(rownames(ins_3))/nrow(ins_3)
scatterplot3d(ins_3[,7:9],color=gray(ins_3$row),xlab = "X",ylab="Y",zlab="Z")
#scatterplot3d(ins_3[,7:9],color="red")

```

###Class 4

```{r}
ins_4<- as.data.frame(cbind(x=t(filter(x,class==4))[,1],y=t(filter(y,class==4))[,1],z=t(filter(z,class==4))[,1]))[-1,]

ins_4<- data.frame(ins_4,vel_x=cumsum(ins_4$x),vel_y=cumsum(ins_4$y),vel_z=cumsum(ins_4$z))
ins_4<- data.frame(ins_4,pos_x=cumsum(ins_4$vel_x),pos_y=cumsum(ins_4$vel_y),pos_z=cumsum(ins_4$vel_z))

rownames(ins_4) <- 1:nrow(ins_4)
ins_4$row<-as.numeric(rownames(ins_4))/nrow(ins_4)
scatterplot3d(ins_4[,7:9],color=gray(ins_4$row),xlab = "X",ylab="Y",zlab="Z")
#scatterplot3d(ins_4[,7:9],color="red")

```

###Class 5

```{r}
ins_5<- as.data.frame(cbind(x=t(filter(x,class==5))[,1],y=t(filter(y,class==5))[,1],z=t(filter(z,class==5))[,1]))[-1,]

ins_5<- data.frame(ins_5,vel_x=cumsum(ins_5$x),vel_y=cumsum(ins_5$y),vel_z=cumsum(ins_5$z))
ins_5<- data.frame(ins_5,pos_x=cumsum(ins_5$vel_x),pos_y=cumsum(ins_5$vel_y),pos_z=cumsum(ins_5$vel_z))

rownames(ins_5) <- 1:nrow(ins_5)
ins_5$row<-as.numeric(rownames(ins_5))/nrow(ins_5)
scatterplot3d(ins_5[,7:9],color=gray(ins_5$row),xlab = "X",ylab="Y",zlab="Z")
#scatterplot3d(ins_5[,7:9],color="red")

```

###Class 6

```{r}
ins_6<- as.data.frame(cbind(x=t(filter(x,class==6))[,3],y=t(filter(y,class==6))[,3],z=t(filter(z,class==6))[,3]))[-1,]

ins_6<- data.frame(ins_6,vel_x=cumsum(ins_6$x),vel_y=cumsum(ins_6$y),vel_z=cumsum(ins_6$z))
ins_6<- data.frame(ins_6,pos_x=cumsum(ins_6$vel_x),pos_y=cumsum(ins_6$vel_y),pos_z=cumsum(ins_6$vel_z))

rownames(ins_6) <- 1:nrow(ins_6)
ins_6$row<-as.numeric(rownames(ins_6))/nrow(ins_6)
scatterplot3d(ins_6[,7:9],color=gray(ins_6$row),xlab = "X",ylab="Y",zlab="Z")
#scatterplot3d(ins_6[,7:9],color="red")

```

###Class 7

```{r}
ins_7<- as.data.frame(cbind(x=t(filter(x,class==7))[,3],y=t(filter(y,class==4))[,3],z=t(filter(z,class==7))[,3]))[-1,]

ins_7<- data.frame(ins_7,vel_x=cumsum(ins_7$x),vel_y=cumsum(ins_7$y),vel_z=cumsum(ins_7$z))
ins_7<- data.frame(ins_7,pos_x=cumsum(ins_7$vel_x),pos_y=cumsum(ins_7$vel_y),pos_z=cumsum(ins_7$vel_z))

rownames(ins_7) <- 1:nrow(ins_7)
ins_7$row<-as.numeric(rownames(ins_7))/nrow(ins_7)
scatterplot3d(ins_7[,7:9],color=gray(ins_7$row),xlab = "X",ylab="Y",zlab="Z")
#scatterplot3d(ins_7[,7:9],color="red")


```

###Class 8

```{r}
ins_8<- as.data.frame(cbind(x=t(filter(x,class==8))[,1],y=t(filter(y,class==8))[,1],z=t(filter(z,class==8))[,1]))[-1,]

ins_8<- data.frame(ins_8,vel_x=cumsum(ins_8$x),vel_y=cumsum(ins_8$y),vel_z=cumsum(ins_8$z))
ins_8<- data.frame(ins_8,pos_x=cumsum(ins_8$vel_x),pos_y=cumsum(ins_8$vel_y),pos_z=cumsum(ins_8$vel_z))

rownames(ins_8) <- 1:nrow(ins_8)
ins_8$row<-as.numeric(rownames(ins_8))/nrow(ins_8)
scatterplot3d(ins_8[,7:9],color=gray(ins_8$row),xlab = "X",ylab="Y",zlab="Z")
#scatterplot3d(ins_8[,7:9],color="red")

```

##TASK 2

###Data Preparation

In order to perform task 2, the data is transformed into long format. 

```{r}
colnames(x)[2:316]<-c(1:315)
colnames(y)[2:316]<-c(1:315)
colnames(z)[2:316]<-c(1:315)

x<-cbind(observation=c(1:nrow(x)),x)
x_long<-gather(x,key="time index", value= "x",c(3:317))

y<-cbind(observation=c(1:nrow(y)),y)
y_long<-gather(y,key="time index", value= "y",c(3:317))

z<-cbind(observation=c(1:nrow(z)),z)
z_long<-gather(z,key="time index", value= "z",c(3:317))

df<-cbind(x_long,y=y_long$y,z=z_long$z)
```

Before doing PCA, minimum and maximum values of the x,y,z values are checked if scaling is needed. According to the PCA results, first component explains 49% of the variance. Composition of the x,y,z in component 1 can be obtained by loadings below.

```{r}
summary(df) #To check if scaling is needed.
df_pca<-princomp(df[,-c(1:3)])
summary(df_pca, loadings=T)

```

```{r}
new_df<-cbind(df[,c(1:3)],mapping_c1=df_pca$scores[,1]) #Mapping of component 1.
head(new_df)
```

The time series plots below, shows the PCA mapping of the first two observations for each class. When each class are observed separately, it can be seen that the two obervations have a very similar trend. When we compare the classes between each other, they are different from each other. However, the differenciation is not enough to class them in 8 groups.

```{r}
class_1_pca<-filter(new_df,class==1)
#Observation 11 and 17 are selected.
class_1_ts<-as.data.frame(cbind(time_index=c(1:315),obs_11=filter(class_1_pca,observation==11)$mapping_c1,obs_17=filter(class_1_pca,observation==17)$mapping_c1))


ggplot(class_1_ts)+
  geom_line(aes(x=time_index,y=obs_11),color="red")+
   geom_line(aes(x=time_index,y=obs_17),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 1")
```


```{r}
class_2_pca<-filter(new_df,class==2)
#Observation 15 and 20 are selected.
class_2_ts<-as.data.frame(cbind(time_index=c(1:315),obs_15=filter(class_2_pca,observation==15)$mapping_c1,obs_20=filter(class_2_pca,observation==20)$mapping_c1))


ggplot(class_2_ts)+
  geom_line(aes(x=time_index,y=obs_15),color="red")+
   geom_line(aes(x=time_index,y=obs_20),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 2")
```


```{r}
class_3_pca<-filter(new_df,class==3)
#Observation 4 and 13 are selected.
class_3_ts<-as.data.frame(cbind(time_index=c(1:315),obs_4=filter(class_3_pca,observation==4)$mapping_c1,obs_13=filter(class_3_pca,observation==13)$mapping_c1))


ggplot(class_3_ts)+
  geom_line(aes(x=time_index,y=obs_4),color="red")+
   geom_line(aes(x=time_index,y=obs_13),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 3")
```


```{r}
class_4_pca<-filter(new_df,class==4)
#Observation 5 and 8 are selected.
class_4_ts<-as.data.frame(cbind(time_index=c(1:315),obs_5=filter(class_4_pca,observation==5)$mapping_c1,obs_8=filter(class_4_pca,observation==8)$mapping_c1))


ggplot(class_4_ts)+
  geom_line(aes(x=time_index,y=obs_5),color="red")+
   geom_line(aes(x=time_index,y=obs_8),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 4")
```


```{r}
class_5_pca<-filter(new_df,class==5)
#Observation 2 and 3 are selected.
class_5_ts<-as.data.frame(cbind(time_index=c(1:315),obs_2=filter(class_5_pca,observation==2)$mapping_c1,obs_3=filter(class_5_pca,observation==3)$mapping_c1))


ggplot(class_5_ts)+
  geom_line(aes(x=time_index,y=obs_2),color="red")+
   geom_line(aes(x=time_index,y=obs_3),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 5")
```


```{r}
class_6_pca<-filter(new_df,class==6)
#Observation 1 and 10 are selected.
class_6_ts<-as.data.frame(cbind(time_index=c(1:315),obs_1=filter(class_6_pca,observation==1)$mapping_c1,obs_10=filter(class_6_pca,observation==10)$mapping_c1))


ggplot(class_6_ts)+
  geom_line(aes(x=time_index,y=obs_1),color="red")+
   geom_line(aes(x=time_index,y=obs_10),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 6")
```


```{r}
class_7_pca<-filter(new_df,class==7)
#Observation 7 and 12 are selected.
class_7_ts<-as.data.frame(cbind(time_index=c(1:315),obs_7=filter(class_7_pca,observation==7)$mapping_c1,obs_12=filter(class_7_pca,observation==12)$mapping_c1))


ggplot(class_7_ts)+
  geom_line(aes(x=time_index,y=obs_7),color="red")+
   geom_line(aes(x=time_index,y=obs_12),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 7")
```


```{r}
class_8_pca<-filter(new_df,class==8)
#Observation 6 and 21 are selected.
class_8_ts<-as.data.frame(cbind(time_index=c(1:315),obs_6=filter(class_8_pca,observation==6)$mapping_c1,obs_21=filter(class_8_pca,observation==21)$mapping_c1))


ggplot(class_8_ts)+
  geom_line(aes(x=time_index,y=obs_6),color="red")+
   geom_line(aes(x=time_index,y=obs_21),color="blue")+
  xlab("Time")+ ylab("Component 1 Mapping")+
  ggtitle("Class 8")
```


##TASK 3
 
In task 3, the data is splitted according to class information and principal component analysis is applied for each class separately. According to results, proportion of the variance explained by the fist component has increased. It is because we have already reduce the noise coming from the difference in class differences.

```{r}
class_based_pca<-df %>% split(.$class) %>% map(~princomp(.[,-c(1:3)]))
map(class_based_pca,summary)
```
 
##TASK 4

According to the plot below, observations having the same class are very close to each other. For example, observations from class 3 are on a band in the right side while observations from class 4 are cumulated on the left. However, the classes are not differentiable when the color codes do not exist.
 
```{r}
t4<-cbind(x[,-c(1:2)],y[,-c(1:2)],z[,-c(1:2)])
t4_m<-as.matrix(t4)
dist_mat<-as.matrix(dist(t4_m, method = "euclidean"))
mds_fit <- as.data.frame(cmdscale(dist_mat,eig=TRUE, k=2)$points[,c(1:2)]) # k is the number of dim
mds<-as.data.frame(cbind(class=x$class,comp_1=mds_fit$V1,comp_2=mds_fit$V2))
mds$class<-as.factor(mds$class)

cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(mds)+
  geom_point(aes(x=comp_1,y=comp_2,color=class))+
  scale_color_manual(values=cbp1)+
  xlab("First Component")+
  ylab("Second Component")
```
 
 